\input{configuration}

\title{Lecture 16 --- Recursive Functions }

\author{J. Zarnett\\
\texttt{jzarnett@uwaterloo.ca}}
\institute{Department of Electrical and Computer Engineering \\
  University of Waterloo}
\date{\today}

\begin{document}

\begin{frame}
  \titlepage
  
 \end{frame}

\begin{frame}
\frametitle{Recursion}

A function that includes a call to itself is said to be \alert{recursive}.

At first glance, this might seem like a really strange idea.\\
\quad Why would a function call itself?

Recall that we said earlier that a common engineering strategy is to break a big problem down into some smaller problems.

Recursion is useful if we are breaking down a problem so that the subproblems are smaller versions of the same problem.

\end{frame}

\begin{frame}
\frametitle{Recursion Example: Factorial}

You may have learned in math class about the factorial.\\
\quad Commonly written $n!$ in mathematical notation for a number $n$.

The factorial of a non-negative integer $n$ is defined as the product of all positive integers less than or equal to $n$.

Examples: \\
\quad $1! = 1$\\
\quad $2! = 2 \times 1 = 2$\\
\quad $3! = 3 \times 2 \times 1 = 6$\\
\quad $4! = 4 \times 3 \times 2 \times 1 = 24$\\
\quad $5! = 5 \times 4 \times 3 \times 2 \times 1 = 120$

\end{frame}

\begin{frame}
\frametitle{Recursion Example: Factorial}

Observe that $5!$ could be rewritten as $5 \times 4!$.

The problem of $5!$ can be broken down into two subproblems:

\begin{enumerate}
	\item $4!$
	\item $5 \times$ the result of (1).
\end{enumerate}

The subproblem $4!$ is a smaller version of the problem we're working on ($5!$), so this problem is a good candidate for recursion.

\end{frame}

\begin{frame}[fragile]
\frametitle{Recursion Example: Factorial}
Suppose now you are going to implement the factorial function.

\begin{verbatim}
static int factorial ( int n )
{
    return n * factorial( n - 1 );
}
\end{verbatim}

There is a problem with this implementation. 

\end{frame}

\begin{frame}
\frametitle{The Factorial Problem}

What happens when the expression n - 1 becomes zero or negative?

This loop will continue forever...\\
\quad Except in practice this will be stopped by an error.

Running this program produces an error called a \alert{Stack Overflow}.

In a later lecture we will discuss what a stack is and how it works.\\
\quad For now, a simplified view of the problem.


\end{frame}

\begin{frame}
\frametitle{Stack Overflow}

When \texttt{Main()} calls \texttt{factorial()}, the computer needs to keep track of where it was in \texttt{Main()} at the time that it went to the other function.

It puts that information in a designated memory area called the stack.

Each time \texttt{factorial()} calls itself, more information is added to the stack to keep track of where it was in \texttt{factorial()}.

If we do this too many times, the stack gets ``full'' (exceeds available memory for it) and this results in a stack overflow error.

\end{frame}

\begin{frame}[fragile]
\frametitle{Factorial Problem}
\begin{verbatim}
static int factorial ( int n )
{
    return n * factorial( n - 1 );
}
\end{verbatim}

The above implementation lacks a \alert{stopping condition}.

The function will keep calling itself until a stack overflow occurs.

\end{frame}

\begin{frame}
\frametitle{Factorial Problem}

Even if program execution did not terminate abnormally as a result of a stack overflow, there's another problem, mathematically.\\
\quad $5! = 5 \times 4 \times 3 \times 2 \times 1$

When we reach 1, we no longer multiply by the next smallest integer.

Our stopping case is therefore when \texttt{n} is 1.

The way the code is written, when \texttt{n} is 1 the evaluation goes on and the expression to the right of \texttt{return} is \texttt{1 * factorial( 0 )}. 

\end{frame}

\begin{frame}[fragile]
\frametitle{Factorial Fixed}
\begin{verbatim}
static int factorial ( int n )
{
    if ( n == 1 )
    {
        return 1;
    }
    return n * factorial( n - 1 );
}
\end{verbatim}

The revised implementation has the stopping condition of \texttt{n} equals 1.

The function will keep calling itself until \texttt{n} is 1.\\
\quad When that happens, 1 is returned.

\end{frame}

\begin{frame}
\frametitle{Recursion}
To make proper use of recursion, we need:
\begin{enumerate}
	\item One or more cases in which the function calls itself; and
	\item One or more cases in which the function does not call itself.
\end{enumerate}

Point (2) is called the stopping case or base case.

As we saw, without a properly defined stopping case, recursion will result in a stack overflow.

\end{frame}

\begin{frame}
\frametitle{Recursion versus Iteration}

Recursion can be a difficult or confusing topic. Is it strictly necessary?

Any task that can be accomplished with recursion can be done without using recursion, such as using a loop (iteratively).

As a small note on efficiency, a recursively written function may run slower than an iteratively written one. Recursion has two advantages:

\begin{enumerate}
	\item Write less code; and \\
	\item The computer tracks state on the stack; iteratively, we must keep track of the state ourselves.
\end{enumerate}

\end{frame}

\begin{frame}[fragile]
\frametitle{Factorial Iteratively}
Here's the factorial function implemented iteratively:

\begin{verbatim}
static int factorial ( int n )
{
    int product = 1;
    for ( int i = n; i > 1; i-- )
    {
        product *= i;
    }
    return product;
}
\end{verbatim}

The iterative implementation uses a \texttt{for} loop, but it could have been written using a \texttt{while} loop.

\end{frame}

% Note: Acknowledge Absolute C++ Book for this in the notes
\begin{frame}
\frametitle{Thinking Recursively}
When designing a recursive function, there are three important criteria to consider:

\begin{enumerate}
	\item There is no infinite recursion;\\
	\quad (A chain of recursive calls eventually reaches a stopping case)
	\item Each stopping case returns the correct value for that case; and
	\item The final returned value is correct if the recursive call(s) returns the correct value(s).
\end{enumerate}

This is an example of \textit{mathematical induction}.\\
If you have not yet learned about it, ignore this for now; you will see it next term in ECE~103 (Discrete Mathematics).

\end{frame}

\begin{frame}
\frametitle{Applying Recursive Thinking}
Consider another mathematical problem, exponentiation: $x^{y}$.

If we wrote a function signature for exponentiation:\\ \texttt{int pow ( int x, int y )}

Is this problem a good candidate for recursive solution?

Mathematically, $a^{b} = a \times a^{b-1}$.

So, yes, this is the kind of problem that lends itself well to recursion.


\end{frame}

\begin{frame}[fragile]
\frametitle{Applying Recursive Thinking}
Now, let's write an implementation for \texttt{pow()}.

\begin{verbatim}
// Precondition: y >= 0
// Postcondition: returns x to the power of y
static int pow ( int x, int y )
{
    if ( y == 0 )
    {
        return 1;
    }
    return pow( x, y - 1 ) * x;
}
\end{verbatim}

\end{frame}

\begin{frame}
\frametitle{Applying the Test}
Let's examine the design criteria and see if this is going to work.

\textit{1. There is no infinite recursion.}\\
\quad The second argument to \texttt{pow(x, y)} is decreased by 1 in each call,\\ \quad so eventually we must get to \texttt{pow(x, 0)}, a stopping case.\\
\quad \quad (As long as the precondition is not violated.)

Thus, there is no infinite recursion; criterion 1 is satisfied.

\end{frame}

\begin{frame}
\frametitle{Applying the Test}

\textit{2. Each stopping case returns the correct value for that case.}\\
\quad Yes. $x^{0} = 1$ is mathematically correct.

\textit{3. The final returned value is correct if the recursive call returns the correct value.}\\
\quad \texttt{pow( x, y - 1 ) * x} follows the rule that $a^{b} = a^{b-1} \times a$.

Criteria 2 and 3 are satisfied.

Having checked those things, we can now be satisfied that the implementation of \texttt{pow()} is correct. 

\end{frame}

\begin{frame}[fragile]
\frametitle{Exponentiation Iteratively}
Here's the exponential function implemented iteratively:

\begin{verbatim}
static int pow ( int x, int y )
{
    int result = 1;
    for ( int i = 0; i < y; i++ )
    {
        result *= x;
    }
    return result;
}
\end{verbatim}

Like the factorial function, the iterative implementation uses a \texttt{for} loop, but it could have been written using a \texttt{while} loop.

\end{frame}

\end{document}

