\input{configuration}

\title{Lecture 19 --- Exception Handling }

\author{J. Zarnett\\
\texttt{jzarnett@uwaterloo.ca}}
\institute{Department of Electrical and Computer Engineering \\
  University of Waterloo}
\date{\today}

\begin{document}

\begin{frame}
  \titlepage
  
  \begin{center}
  \small{Acknowledgments: W.D. Bishop}
  \end{center}
\end{frame}

\begin{frame}
\frametitle{Kinds of Exceptions}

An exception is an instance of a detected run-time error.

Exceptions can be generated ``automatically'' or thrown ``manually''.

A ``manually'' generated exception is when the keyword \texttt{throw} appears in the code. A programmer has explicitly \texttt{throw}n an exception.

An ``automatically'' generated exception appears without the appearance of the \texttt{throw} keyword. An example is a division by zero.

The line of code is\\
\quad \texttt{double result = input / divisor; }\\
 but an exception will occur here if divisor is zero.

\end{frame}


\begin{frame}
\frametitle{Exceptions are Thrown}


In reality, exceptions are always thrown... 

Some exceptions are thrown by the system libraries and built-in code, so they appear to be ``automatically'' generated.

Regardless of the source, handling is the same.

\end{frame}


\begin{frame}
\frametitle{Exception Handling}

The process of detecting an exception and attempting to correct the situation is known as exception handling.

Exceptions can be handled or unhandled:
\begin{itemize}
	\item A \alert{handled} exception is caught by a portion of the program and corrective action is taken to fix the situation
	\item An \alert{unhandled} exception is not caught by a portion of the program and it results in the termination of the program
\end{itemize}

\end{frame}

\begin{frame}[fragile]
\frametitle{Undhandled Exception}

Here is an example of an unhandled exception:

\begin{verbatim}
static void Main( )
{
    int i = 0;		
    Console.Write( "100 / 0 = " );
    Console.WriteLine ( 100 / i );
}
\end{verbatim}

The C\# compiler is clever enough to detect an error if we explicitly write (100 / 0) as an expression, so we have to put 0 in a variable \texttt{i}.

\end{frame}


\begin{frame}
\frametitle{Handling an Exception}

The keyword for handling an exception is \texttt{catch}.\\
\quad (The opposite of \texttt{throw}.)

Much like someone who is trying to catch a ball in the air, the \texttt{catch} statement needs to be looking in the right direction.

A \texttt{catch} statement therefore is explicitly applied to a block of code.

\end{frame}

\begin{frame}[fragile]
\frametitle{Try-Catch}
The syntax for exception handling is the \texttt{try-catch} statement.

\begin{verbatim}
try
{
    // Statement Block
}
catch( ExceptionType e )
{
    // Exception Handling Block
}    
\end{verbatim}

The keyword \texttt{try} indicates what a \texttt{catch} is looking at.

The \texttt{catch} applies to the statements inside the \texttt{try} block.\\
\quad It will catch an exception only if it comes from within the \texttt{try} block.

\end{frame}

\begin{frame}
\frametitle{ExceptionType}

In the syntax for \texttt{catch} there is: \texttt{ExceptionType e}.

\texttt{Exception} is the general form. There are more specific exception types, to allow us to tell different sorts of errors apart.

In addition to saying where we are looking for an exception to be thrown, we also specify what kind of exception we want to \texttt{catch}.

\texttt{Exception} is general and if we just write \texttt{catch ( Exception e )} we will catch most types of exception.

\end{frame}

\begin{frame}
\frametitle{C\# Important Exception Types}

Some common Exception Types in C\#:

{\scriptsize
\begin{center}
\begin{tabular}{l|l}
\textbf{Exception Name} & \textbf{Description} \\ \hline
\texttt{Exception} & General exception \\ \hline
\texttt{ArgumentException} & Error in an argument passed to a function \\ \hline
\texttt{ArithmeticException} & Some type of arithmetic (mathematical) error \\ \hline
\texttt{DivideByZeroException} & Division by zero arithmetic error \\ \hline
\texttt{ArrayTypeMismatchException} & Type of an array does not match with the type required \\ \hline
\texttt{FormatException} & Problem with the formatting of input \\ \hline
\texttt{IndexOutOfRangeException} & An array index beyond the bounds of an array \\ \hline
\texttt{NullReferenceException} & Use of a null reference \\
\end{tabular}
\end{center}
}

Don't memorize this list, but you may use it for reference.

\end{frame}


\begin{frame}[fragile]
\frametitle{Try-Catch Example}

Let's fill in the try-catch example from before.

\begin{verbatim}
try
{
  double result = input / divisor;
  Console.Write( "Result: ");
  Console.WriteLine( result );
}
catch( DivideByZeroException e )
{
  Console.WriteLine( "Division By Zero Error Detected!" );
}    
\end{verbatim}

\end{frame}

\begin{frame}
\frametitle{Try-Catch Execution}

The only output was the error message.\\
\quad The statements to output ``Result: '' and the number did not occur.

Like \texttt{return} or \texttt{break}, when an exception is encountered, execution of the current block is stopped at the point of the exception.

If an exception occurs, no further statements of the \texttt{try} block run.

The system will then look for a \texttt{catch} block matching the type of the exception that was encountered.

In the example above, there is a matching \texttt{catch} block.\\
\quad What happens if none is found?

\end{frame}

\begin{frame}[fragile]
\frametitle{Looking for Someone to Catch}

The system will continue looking by going up a level to the function that called the function where the exception was encountered.

\begin{verbatim}
static void Main ( )
{
    try
    {
      divideNumbers( 100, 0 );
    }
    catch ( DivideByZeroException e )
    {
      Console.WriteLine( "Division By Zero Detected" );
    }
}
\end{verbatim}

In the above example, an exception happened in \texttt{divideNumbers()} but the \texttt{catch} block is found in \texttt{Main}.

\end{frame}

\begin{frame}
\frametitle{Looking for Someone to Catch}

If the calling function lacks a matching \texttt{catch} block, the system goes up another level and repeats this procedure until it gets to \texttt{Main}.

If \texttt{Main} does not have one either, it's an unhandled exception.\\
\quad The program terminates with an error message.

In both of the preceding examples, we found a \texttt{catch} block matching the type of the exception, so it is handled.

\end{frame}

\begin{frame}[fragile]
\frametitle{Try-Catch Example}
{\small
\begin{verbatim}
try
{
    double result = input / divisor;
    Console.Write( "Result: ");
    Console.WriteLine( result );
}
catch( DivideByZeroException e )
{
  Console.WriteLine( "Division By Zero Error Detected!" );
}    
\end{verbatim}
}

So after the exception occurs, control goes to the \texttt{catch} block and the statement there executes.

Then the next statement executed is after the end of the \texttt{catch} block.\\
\quad Not the console writes.
\end{frame}


\begin{frame}
\frametitle{Variable Scope in Try-Catch}

Note that our usual rules of variable scope apply in a \texttt{try-catch}.

In the previous example \texttt{double result} is only in scope within the \texttt{try} block and not after the \texttt{try-catch} is done.

Like a function parameter, \texttt{DivideByZeroException e} is a parameter available in the \texttt{catch} block.

Right now, we won't make use of \texttt{e}, but if you are feeling ambitious you can play around with what it contains.

\end{frame}

\begin{frame}
\frametitle{Multiple Types of Exceptions}

It is possible to \texttt{catch} more than one type of exceptions from a single \texttt{try} block.

The syntax and execution for this works like an \texttt{if-else} statement.

When an exception is encountered, the exceptions are checked in order from top to bottom, like the \texttt{if-else}.\\
\quad The first \texttt{catch} that matches executes.

Only one of the \texttt{catch} blocks will execute (mutual exclusivity).

\end{frame}

\begin{frame}[fragile]
\frametitle{Multiple Types of Exceptions}


\begin{verbatim}
try
{
   complicatedFunction( );
}
catch( DivideByZeroException e )
{
  Console.WriteLine( "Division By Zero Error Detected!" );
}
catch ( ArithmeticException e2 )
{
  Console.WriteLine( "Arithmetic Error Detected!" );
}
catch ( IndexOutOfRangeException e3 )
{
  Console.WriteLine( "Index out of range!" );
}
\end{verbatim}


\end{frame}

\begin{frame}[fragile]
\frametitle{Handling All Exceptions}

It's possible to catch most exceptions using \texttt{catch ( Exception e )}

However, there are a few exceptions that will not be caught by this.

To be sure to \texttt{catch} \underline{all} exceptions, use the \texttt{catch} statement without the brackets and exception type:

\begin{verbatim}
try
{
    // Try block
}
catch
{
    // All exceptions caught
}
\end{verbatim}

\end{frame}

\begin{frame}
\frametitle{\texttt{finally}}

The \texttt{finally} statement provides a mechanism for executing code immediately after the (partial) execution of a try block.

It may be desirable to save the state of the program prior to exiting, regardless of whether an exception has been handled or not.

The \texttt{finally} statement enables a program to exit gracefully, even in the presence of exceptions.

The \texttt{finally} block comes after the \texttt{catch} blocks, if any.

The \texttt{finally} block executes regardless of whether or not an exception is thrown.

\end{frame}


\begin{frame}[fragile]
\frametitle{\texttt{finally} Syntax}

Here is the syntax for the \texttt{try-catch-finally} statement:

\begin{verbatim}
try
{
    // Statement Block
}
catch( ExceptionType e )
{
    // Exception Handling Block
}
finally
{
    // Finally Block
}
\end{verbatim}

\end{frame}

\begin{frame}[fragile]
\frametitle{\texttt{finally} Syntax}

Here is an example of the \texttt{try-catch-finally} statement:

\begin{verbatim}
try
{
    Console.WriteLine( "Attempting Division..." );
    divideNumbers(100, 0);
    Console.WriteLine( "Division Succeeded." );
}
catch( ExceptionType e )
{
    Console.WriteLine( "Division by zero detected." );
}
finally
{
    Console.WriteLine( "Operation Complete." );
}
\end{verbatim}

\end{frame}

\begin{frame}[fragile]
\frametitle{\texttt{finally} Syntax}

What if we don't have \texttt{catch}?

\begin{verbatim}
try
{
    Console.WriteLine( "Attempting Division..." );
    divideNumbers(100, 0);
    Console.WriteLine( "Division Succeeded." );
}
finally
{
    Console.WriteLine( "Operation Complete." );
}
\end{verbatim}

The \texttt{finally} block runs before the exception terminates the program.

\end{frame}



\begin{frame}
\frametitle{What to Do with Exceptions}

Sometimes, when an exception is encountered, the program can take some corrective action and deal with the problem in the \texttt{catch} block.

In this lecture we have mostly handled errors by simply reporting them and carrying on.

If the program is a calculator and it takes as input two numbers and an operation, one exception should not end the program.

In other cases, carrying on is harmful and we should stop execution of the program before anything else goes wrong.

\end{frame}




\end{document}

